---
title: "Senescence-Driven Cardiac Fibrosis After Myocardial Infarction — Target Discovery for Anti-Fibrotic Gene Therapy"
author: "Tehrim Naveed"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Background and Question

Cardiac fibroblast senescence after myocardial infarction (MI) appears to be
time-dependent: transient senescence in the first ~3-7 days post-MI limits
excessive myofibroblast proliferation and helps prevent ventricular rupture,
but if senescent fibroblasts persist, they drive chronic SASP-mediated
fibrosis and impair long-term cardiac function. The transcriptional switch
between these two states, reparative and transient versus persistent and
pathological, is not well defined at single-cell resolution in human tissue.

**Question:** What distinguishes the transcriptional program of transiently
senescent, reparative cardiac fibroblasts from persistently senescent,
SASP-high fibroblasts after MI, and which of the genes driving that
divergence are plausible targets for AAV- or siRNA-based anti-fibrotic gene
therapy?

**Dataset:** Kuppe et al. 2022, *Nature* (DOI: 10.1038/s41586-022-05060-x) —
single-nucleus RNA-seq, snATAC-seq, and spatial transcriptomics (Visium)
across ischemic, border, remote, and control zones of human MI tissue. This
analysis uses the snRNA-seq component (191,795 nuclei, n = 23 samples).

---

# 2. Literature Review

## Hernández et al., *Scientific Data* 2026
Defines a Reparative Cardiac Fibroblast (RCF) subpopulation responsible for
orchestrating scar formation and preventing ventricular rupture post-MI.
Combines bulk RNA-seq, RNAscope, and spatial transcriptomics to map the
early transcriptional transition into the RCF state. Notably, the timing of
RCF appearance and the specific molecular events driving the transition are
described as still undefined, this is the most direct precedent for the
gap this project addresses, and I'll use it to cross-check any early
reparative markers recovered from the Kuppe fibroblast clusters.

## Frontiers in Immunology 2026 — SASP-mediated senescence post-MI (review)
Synthesizes the current model: fibroblast proliferation peaks around day 7
post-MI, and transient, controlled senescence at this stage limits excessive
scarring. Under persistent inflammatory/pro-fibrotic signaling, senescence
shifts from reparative regulation to pathological amplification via
long-term SASP secretion. Provides a working list of senescence/SASP
therapeutic strategies (senolytics, senostatics) to compare against
gene-therapy framing.

## Cardiac fibroblast senescence dynamics — bioRxiv, Aug 2025
Multi-omic (spatial + single-nucleus) profiling in a p16 (Cdkn2a) senescence-
reporter mouse model, day 7 post-MI. Fibroblasts account for the majority
(~58%) of p16-high senescent cells across all cardiac cell types. Senescent
cell burden is shown to be transient, most p16-high cells revert to a
non-senescent state over the following weeks, supporting a dynamic rather
than fixed senescent state. Strongest mechanistic precedent so far for
treating senescence as a reversible, time-resolved process rather than a
binary label.

## Kuppe et al. 2022, *Nature* — primary dataset
Defines fibroblast-to-myofibroblast differentiation trajectories across
injury zones and validates disease-specific cardiac cell states in their
spatial context. Fibrotic zone (FZ) and ischemic zone (IZ) samples are the
most relevant for capturing late/persistent fibrotic states; border zone
(BZ) and remote zone (RZ) samples should capture earlier or less-affected
transcriptional states. No explicit senescence/SASP scoring is reported in
the original paper, this is the gap this reanalysis is positioned to fill.

---

# 3. Environment Setup

```{r renv-init, eval=FALSE}
# Run once from the console, from the project root
# install.packages("renv")
# renv::init()
```

```{r install-packages, eval=FALSE}
install.packages(c("Seurat", "tidyverse", "patchwork", "Matrix"))

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c(
  "zellkonverter",
  "SingleCellExperiment",
  "DESeq2",
  "clusterProfiler",
  "org.Hs.eg.db",
  "AUCell",
  "slingshot"
))

install.packages("harmony")
install.packages("msigdbr")
# devtools::install_github("jinworks/CellChat")  # Day 5
```

```{r load-libraries}
library(Seurat)
library(tidyverse)
library(patchwork)
library(zellkonverter)
library(SingleCellExperiment)
library(harmony)
```

---

# 4. Data Acquisition

snRNA-seq data (191,795 nuclei, all zones and controls) downloaded from CZ
CELLxGENE Discover, Kuppe et al. 2022 collection, as `.h5ad`, converted to a
Seurat object below.

```{r convert-h5ad, eval=FALSE}
sce <- readH5AD("data/raw/kuppe_2022_scRNA.h5ad")
seurat_obj <- as.Seurat(sce, counts = "X", data = NULL)

seurat_obj
dim(seurat_obj)
head(seurat_obj@meta.data)

saveRDS(seurat_obj, "data/processed/kuppe_seurat_raw.rds")
```

---

# 5. Repository Structure

```
cardiac-mi-senescence/
├── data/
│   ├── raw/
│   └── processed/
├── notebooks/
│   ├── 01_setup_and_literature.Rmd
│   ├── 02_qc_preprocessing.Rmd
│   ├── 03_clustering_annotation.Rmd
│   ├── 04_differential_expression.Rmd
│   └── 05_pathway_enrichment.Rmd
├── results/
│   ├── figures/
│   └── tables/
├── renv.lock
├── .gitignore
└── README.md
```

---

# 6. Session Info

```{r session-info}
sessionInfo()
```
