---
title: "Senescence-Driven Cardiac Fibrosis After MI — Gene Therapy Target Discovery"
subtitle: "Day 1: Setup, Data Acquisition, and Literature Notes"
author: "Tehrim [Your Last Name]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Project Question

**Working hypothesis:** Cardiac fibroblast senescence after myocardial infarction (MI)
is beneficial when transient (limits over-fibrosis, days ~3-7 post-MI) but becomes
pathological when persistent (drives chronic SASP-mediated fibrosis). The molecular
switch-point between these two states is poorly defined and is the gap this project
targets — with an eye toward identifying transcriptional drivers that could be
AAV/siRNA gene-therapy targets to "lock in" the reparative state and prevent the
pathological one.

**Dataset:** Kuppe et al. 2022, *Nature* — spatial multi-omic map of human MI
(single-cell RNA-seq + spatial transcriptomics across injury zones and controls).
DOI: 10.1038/s41586-022-05060-x

---

# 2. Literature Notes

*Fill this in as you read today — this section becomes your Introduction later.*

## Hernández et al., Sci Data 2026 — Reparative Cardiac Fibroblasts (RCFs)
- Key claim:
- Relevance to my question:
- Any gene markers they report for the reparative state:

## Frontiers Immunol. 2026 — SASP-mediated senescence post-MI (review)
- Key claim:
- Relevance to my question:

## biorxiv Aug 2025 — p16 reporter mouse, multi-omic senescence MI
- Key claim:
- Relevance to my question:
- Any gene markers/timepoints they report:

## Kuppe et al. 2022, Nature — dataset paper
- Fibroblast states they define:
- Spatial zones in the data (ischemic / border / remote / control):
- Anything about senescence or SASP genes already flagged:

---

# 3. Environment Setup

This project uses `renv` for reproducibility — this is what makes the notebook
shareable and rerunnable by anyone (including a PI reviewing your GitHub).

```{r renv-init, eval=FALSE}
# Run this once, in the R console (not inside the notebook), from your project folder:
# install.packages("renv")
# renv::init()
```

```{r install-packages, eval=FALSE}
# Core single-cell workflow
install.packages(c("Seurat", "tidyverse", "patchwork", "Matrix"))

# Bioconductor packages
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c(
  "zellkonverter",   # reads .h5ad (AnnData) files into SingleCellExperiment
  "SingleCellExperiment",
  "DESeq2",
  "clusterProfiler",
  "org.Hs.eg.db",
  "AUCell",          # senescence/SASP gene set scoring
  "slingshot"        # trajectory analysis
)

# Batch correction
install.packages("harmony")

# Gene set database
install.packages("msigdbr")

# Cell-cell communication (Day 5, optional)
# devtools::install_github("jinworks/CellChat")
```

```{r load-libraries}
library(Seurat)
library(tidyverse)
library(patchwork)
library(zellkonverter)
library(SingleCellExperiment)
library(harmony)
```

---

# 4. Data Acquisition

The Kuppe et al. 2022 data is distributed as an `.h5ad` (AnnData/Python format).
We need to convert it to a Seurat object.

**Manual download step (do this outside R):**

1. Go to the CZ CELLxGENE Discover collection page for "Spatial multi-omic map of
   human myocardial infarction" (search the title at cellxgene.cziscience.com).
2. Download the single-cell RNA-seq `.h5ad` file (not the spatial one, for now —
   we start with the scRNA-seq to keep Day 1-4 manageable).
3. Save it into `data/raw/kuppe_2022_scRNA.h5ad` in your project folder.

```{r convert-h5ad, eval=FALSE}
# Convert .h5ad -> SingleCellExperiment -> Seurat
sce <- readH5AD("data/raw/kuppe_2022_scRNA.h5ad")
seurat_obj <- as.Seurat(sce, counts = "X", data = NULL)

# Quick sanity check
seurat_obj
dim(seurat_obj)
head(seurat_obj@meta.data)

# Save so you don't have to reconvert every time
saveRDS(seurat_obj, "data/processed/kuppe_seurat_raw.rds")
```

---

# 5. Project Repo Structure

Create this folder structure in your project directory (same logic as a typical
GitHub-shared bioinformatics repo — this is what a PI or collaborator expects
to see):

```
cardiac-mi-senescence/
├── data/
│   ├── raw/            # original .h5ad (add to .gitignore — too large for GitHub)
│   └── processed/       # saved .rds Seurat objects
├── notebooks/
│   ├── 01_setup_and_literature.Rmd    <- this file
│   ├── 02_qc_preprocessing.Rmd
│   ├── 03_clustering_annotation.Rmd
│   ├── 04_differential_expression.Rmd
│   └── 05_pathway_enrichment.Rmd
├── results/
│   ├── figures/
│   └── tables/
├── renv.lock            # auto-generated — locks exact package versions
├── .gitignore
└── README.md
```

`.gitignore` should contain at minimum:
```
data/raw/
*.rds
.Rproj.user/
.Rhistory
```

---

# 6. Session Info (reproducibility — always keep this at the end of every notebook)

```{r session-info}
sessionInfo()
```
