---
title: "Senescence Driven Cardiac Fibrosis After Myocardial Infarction, Target Discovery for Anti Fibrotic Gene Therapy"
subtitle: "Day 3: Clustering, Ambient RNA Correction, Senescence Scoring, and Annotation"
author: "Tehrim Naveed"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load-libraries}
library(Seurat)
library(tidyverse)
library(harmony)
library(SingleCellExperiment)
library(zellkonverter)
library(org.Hs.eg.db)
library(msigdbr)
library(AUCell)
library(celda)
library(scDblFinder)
```

# 1. Objective

Clustered and annotated the 46,453 fibroblast nuclei quality
controlled in Day 2, corrected for ambient RNA contamination flagged as a
known limitation in the original publication, called score for each fibroblast for senescence using two independent methods, identify what each resulting cluster represents using marker genes, and validate my cell population with an independent doublet detection pass.

---

# 2. Doublet Validation on the Full Dataset

Since doublets are only reliably detectable when more than one cell type is
present to simulate against, I ran doublet detection before subsetting to
fibroblasts, on the full, all cell type object, using cell_type_original as
the clustering variable and donor_id as the sample grouping variable so that
detection respects real physical capture boundaries.

```{r full-dataset-doublet-check, eval=FALSE}
sce_full <- readH5AD("data/raw/kuppe_2022_scRNA.h5ad")
counts_full <- assay(sce_full, "X")
meta_full <- as.data.frame(colData(sce_full))
rownames(meta_full) <- colnames(counts_full)
rm(sce_full)
gc()

sce_input_full <- SingleCellExperiment(assays = list(counts = counts_full))

sce_input_full <- scDblFinder(
  sce_input_full,
  clusters = meta_full$cell_type_original,
  samples = meta_full$donor_id
)

meta_full$our_doublet_class <- sce_input_full$scDblFinder.class

table(meta_full$our_doublet_class, meta_full$doublet_score < 0.3)
```

This identified 4,438 likely doublets across the full dataset, close to the
4,276 cells flagged by the original authors' own scDblFinder run
(doublet_score < 0.3), confirming that my independent run recovers a similar
overall doublet rate to the published one. Individual cell level doublet
calls between two independently run instances of this method are known to
vary due to its reliance on randomized simulation, so rather than trust
either run in isolation, I treated the intersection of both calls as a high
confidence doublet set, and cells uniquely flagged by only one run as a
lower confidence signal worth tracking rather than acting on directly.

---

# 3. Load Fibroblast Data

```{r load-data}
fibroblasts <- readRDS("data/processed/kuppe_fibroblasts_qc.rds")
dim(fibroblasts)
```

This object already reflects the doublet_score < 0.3 filter applied in Day
2, using the original authors' own scDblFinder output.

---

# 4. Normalization, Variable Features, and PCA

```{r normalize-variable-features}
fibroblasts <- NormalizeData(fibroblasts)
fibroblasts <- FindVariableFeatures(fibroblasts, selection.method = "vst", nfeatures = 2000)
```

I used the vst method since it selects genes based on variance relative to
expression level, preventing highly expressed genes from dominating
selection purely due to scale. Among the top variable genes were SERPINE1,
POSTN, and FN1, established markers of myofibroblast activity and the
senescence associated secretory phenotype, an early signal that this
fibroblast population carries the biology this project is built around.

Two of the top ten variable genes, ENSG00000285861 and ENSG00000286922,
mapped to long non coding RNAs currently annotated only as "novel
transcript" with no established function, confirmed directly against the
Ensembl REST API. I retained these in the analysis rather than excluding
them, since their role is currently unknown rather than ruled out, and did
not filter non coding genes from the variable feature set at any point.

```{r scale-pca}
fibroblasts <- ScaleData(fibroblasts, features = VariableFeatures(fibroblasts))
fibroblasts <- RunPCA(fibroblasts, features = VariableFeatures(fibroblasts))

ElbowPlot(fibroblasts, ndims = 30)
```

I selected fifteen principal components for downstream steps, based on the
elbow plot flattening substantially by around PC10 to PC12.

---

# 5. Ambient RNA Correction

Kuppe et al. 2022 report that necrotic tissue in the ischemic zone releases
ambient RNA that can contaminate neighboring nuclei during capture, and flag
this as a limitation in their own analysis that was not computationally
corrected. Given this, I applied DecontX, which estimates and removes
ambient contamination directly from the filtered count matrix, appropriate
here since the raw, unfiltered droplet matrix needed for tools like SoupX or
CellBender is not available through this dataset's public distribution.

```{r initial-cluster-for-decontx}
fibroblasts <- RunHarmony(
  fibroblasts,
  group.by.vars = "donor_id",
  reduction.use = "pca",
  dims.use = 1:15
)

fibroblasts <- FindNeighbors(fibroblasts, reduction = "harmony", dims = 1:15)
fibroblasts <- FindClusters(fibroblasts, resolution = 0.5)
```

I used this initial clustering to inform DecontX's contamination model,
which estimates each cell's deviation from its assigned cluster's expected
expression profile.

```{r run-decontx}
counts_matrix <- GetAssayData(fibroblasts, layer = "counts")

decontx_results <- decontX(counts_matrix, z = fibroblasts@meta.data$seurat_clusters)

fibroblasts@meta.data$contamination <- decontx_results$contamination

decontx_counts <- decontx_results$decontXcounts

fibroblasts[["decontX"]] <- CreateAssay5Object(counts = decontx_counts)
DefaultAssay(fibroblasts) <- "decontX"
```

I added the decontaminated counts as a separate assay rather than
overwriting the original counts, keeping both available for comparison. By
running DecontX before finalizing clusters, I address ambient contamination
directly at the source, going a step further than the original publication,
which flagged the issue but did not apply a computational correction.

---

# 6. Final Clustering

```{r decontx-normalize-cluster}
fibroblasts <- NormalizeData(fibroblasts)
fibroblasts <- FindVariableFeatures(fibroblasts, selection.method = "vst", nfeatures = 2000)
fibroblasts <- ScaleData(fibroblasts, features = VariableFeatures(fibroblasts))
fibroblasts <- RunPCA(fibroblasts, features = VariableFeatures(fibroblasts))

fibroblasts <- RunHarmony(
  fibroblasts,
  group.by.vars = "donor_id",
  reduction.use = "pca",
  dims.use = 1:15,
  reduction.save = "harmony_decontx"
)

fibroblasts <- RunUMAP(fibroblasts, reduction = "harmony_decontx", dims = 1:15, reduction.name = "umap_decontx")
fibroblasts <- FindNeighbors(fibroblasts, reduction = "harmony_decontx", dims = 1:15)
fibroblasts <- FindClusters(fibroblasts, resolution = 0.5)

DimPlot(fibroblasts, reduction = "umap_decontx", group.by = "seurat_clusters", label = TRUE)
DimPlot(fibroblasts, reduction = "umap_decontx", group.by = "major_labl")
```

This produced nine clusters on the ambient RNA corrected data. The
ischemic zone remained clearly separated from the other four zones even
after this correction, indicating a genuine transcriptional difference in
IZ fibroblasts, consistent with the necrosis associated stress response
described in the original publication, rather than a technical artifact.

---

# 7. Senescence Scoring

I scored senescence using the SenMayo gene set (Saul et al. 2022), pulled
directly from MSigDB for reproducibility.

```{r senmayo-retrieval}
senmayo <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CGP") %>%
  filter(gs_name == "SAUL_SEN_MAYO")

senmayo_ids <- unique(na.omit(senmayo$ensembl_gene))
senmayo_in_data <- senmayo_ids[senmayo_ids %in% rownames(fibroblasts)]

length(senmayo_ids)
length(senmayo_in_data)
```

124 of the 125 published SenMayo genes were available in MSigDB's current
build, and 116 of those were present as measured features in this dataset.

```{r addmodulescore}
fibroblasts <- AddModuleScore(
  fibroblasts,
  features = list(senmayo_in_data),
  name = "SenMayo"
)
```

```{r aucell}
expr_matrix <- GetAssayData(fibroblasts, layer = "counts")
gene_sets <- list(SenMayo = senmayo_in_data)

cell_rankings <- AUCell_buildRankings(expr_matrix, plotStats = TRUE)
cell_auc <- AUCell_calcAUC(gene_sets, cell_rankings)

fibroblasts@meta.data$SenMayo_AUCell <- as.numeric(getAUC(cell_auc)["SenMayo", ])
```

I scored senescence with two independent methods. AddModuleScore compares
average expression against a matched control gene set, while AUCell ranks
genes within each cell rather than comparing expression magnitude directly,
making it less sensitive to depth differences between cells.

```{r score-correlation}
cor(fibroblasts@meta.data$SenMayo1, fibroblasts@meta.data$SenMayo_AUCell, method = "spearman")
```

I used Spearman correlation because the two scores sit on fundamentally
different scales, AddModuleScore is zero centered and can be negative,
AUCell is bounded between 0 and 1, so comparing relative ordering rather
than raw magnitude is the appropriate test. The two methods showed a
Spearman correlation of 0.87, strong agreement despite being built on
independent statistical logic, which gives me confidence this senescence
signal reflects real biology rather than an artifact of either method's
assumptions.

```{r score-visualization}
VlnPlot(fibroblasts, features = c("SenMayo1", "SenMayo_AUCell"), group.by = "seurat_clusters", pt.size = 0)
VlnPlot(fibroblasts, features = c("SenMayo1", "SenMayo_AUCell"), group.by = "major_labl", pt.size = 0)
```

Both scores were highest in clusters 1, 7, and 8, and highest overall in the
IZ zone. I annotated each cluster by marker genes next, to establish which
of these are genuine fibroblast populations before drawing conclusions from
the scores.

---

# 8. Cluster Annotation

```{r find-markers}
markers <- FindAllMarkers(fibroblasts, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

top_markers <- markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10)

top_markers$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys = top_markers$gene,
  keytype = "ENSEMBL",
  column = "SYMBOL"
)
```

Based on top marker genes, I identified the following cluster identities:

Cluster 0, quiescent fibroblasts: ACSM1, ACSM3, ADH1B, FMO2, metabolic and
detoxification genes consistent with homeostatic, baseline fibroblasts.

Cluster 2, quiescent fibroblasts: SCARA5, PCOLCE2, established markers of
resident cardiac fibroblasts.

Cluster 4, activated myofibroblasts: MEOX1, THBS4, FAP, POSTN, TNC, LOXL1,
the matrix producing, disease relevant fibroblast state central to this
project.

Cluster 5, acutely stressed fibroblasts: FOSB, NR4A1, PER2, immediate early
genes, enriched in the fibrotic zone.

Cluster 1, ischemic zone associated fibroblasts, elevated senescence score:
this cluster carries a metabolic and stress associated marker profile
consistent with fibroblast identity, and is the clearest senescence signal
in a genuine fibroblast population in this dataset.

```{r dotplot-verification}
canonical_ids <- mapIds(
  org.Hs.eg.db,
  keys = c("PTPRC", "CD68", "MKI67", "COL1A1"),
  keytype = "SYMBOL",
  column = "ENSEMBL"
)

DotPlot(fibroblasts, features = canonical_ids, group.by = "seurat_clusters") + RotatedAxis()
```

Clusters 7 and 8, by contrast, expressed canonical macrophage markers
(CD163, MSR1, STAB1, MERTK, PTPRC, CD68) and cell cycle genes (MKI67,
FANCA, BRIP1, MELK, UHRF1) respectively, alongside detectable COL1A1,
consistent with fibroblast-immune and fibroblast-proliferating cell
doublets rather than pure fibroblast populations. I excluded both from
downstream fibroblast specific analysis on this basis, standard practice
when marker evidence contradicts an original cell type label.

Cluster 3 shows some cardiomyocyte marker expression (TNNI3, MYL2, MYH7)
alongside fibroblast genes, which I am carrying forward as a specific point
to investigate further in Day 4, rather than excluding outright given the
less clear cut evidence compared to clusters 7 and 8.

```{r remove-clusters}
fibroblasts_clean <- subset(
  fibroblasts,
  subset = seurat_clusters %in% c("7", "8"),
  invert = TRUE
)

dim(fibroblasts_clean)
table(fibroblasts_clean@meta.data$seurat_clusters, fibroblasts_clean@meta.data$major_labl)
```

This leaves 44,125 fibroblasts across clusters 0 through 6, forming the
population I carry into Day 4.

I also checked this cleaned population against the full dataset doublet
detection results from Section 2. 533 fibroblasts (1.2 percent) were flagged
by my independent run but not the original filter, mildly enriched in
cluster 1 (2.6 percent versus 0.6 to 1.5 percent in other clusters). Given
the known variability in individual doublet calls between independent runs
described earlier, I did not remove these cells, and instead note this as
context for interpreting cluster 1 specifically, alongside its comparatively
lower sequencing depth.

```{r additional-doublet-check, eval=FALSE}
additional_doublets <- rownames(meta_full)[
  meta_full$our_doublet_class == "doublet" & meta_full$doublet_score < 0.3
]

fibroblasts_clean@meta.data$additional_flag <- colnames(fibroblasts_clean) %in% additional_doublets

table(fibroblasts_clean@meta.data$additional_flag, fibroblasts_clean@meta.data$seurat_clusters)
```

---

# 9. Save Output

```{r save-output}
saveRDS(fibroblasts_clean, "data/processed/kuppe_fibroblasts_final.rds")
```

---

# 10. Summary

I corrected for ambient RNA contamination flagged in the original
publication using DecontX, applied before finalizing clusters, which
confirmed the ischemic zone's transcriptional distinctness reflects real
biology rather than a technical artifact. I scored senescence using two
independent methods with strong agreement (Spearman rho = 0.87), then used
marker gene analysis to confirm cluster identities, which allowed me to
exclude two clusters that were immune and proliferating cell populations
rather than fibroblasts. The resulting population, cluster 1 specifically,
represents a fibroblast population with a genuine, corroborated senescence
signal, interpreted alongside its comparatively lower sequencing depth and
a modest doublet enrichment flagged by independent validation. Cluster 3's
partial cardiomyocyte marker expression is carried forward as a specific
point for further investigation in Day 4.

---

# 11. Session Info

```{r session-info}
sessionInfo()
```
