---
title: "Senescence Driven Cardiac Fibrosis After Myocardial Infarction, Target Discovery for Anti Fibrotic Gene Therapy"
subtitle: "Day 4: Differential Expression, Pathway Enrichment, and Target Shortlisting"
author: "Tehrim Naveed"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load-libraries}
library(Seurat)
library(tidyverse)
library(DESeq2)
library(org.Hs.eg.db)
library(msigdbr)
library(clusterProfiler)
library(httr)
library(jsonlite)
```

# 1. Objective

My Day 1 literature review identified a specific question as to how cardiac fibroblast senescence after myocardial infarction appears to be time dependent, transiently reparative in the days immediately following injury, but capable of becoming persistent and pathological if it does not resolve. Hernandez et al. (Scientific Data, 2026) name the early, beneficial state Reparative Cardiac Fibroblasts and explicitly note that the molecular switch point separating it from a persistent, pathological state is not yet defined. A companion bioRxiv study using a p16 senescence reporter mouse supports this framing directly, showing that senescent fibroblast burden after MI is dynamic and substantially reversible over time rather than fixed.

In this notebook, I use the fibroblast population established and validated
on Day 3, 39,256 nuclei with ambient RNA correction, marker verified cluster
identity, and doublet cross validation, to directly compare fibroblasts with
high versus low senescence signal and identify the genes that distinguish
them. My goal is to shortlist evidence backed list of candidate genes for antifibrotic gene therapy, genes whose behavior could plausibly be modulated to favor the reparative state and limit the pathological one.

---

# 2. Defining Senescence Groups

Both senescence scores computed on Day 3, AddModuleScore and AUCell, are
continuous. To run a differential expression comparison I converted them into discrete high and low groups, using a consensus approach that requires both independently computed scores to agree, rather than relying on either method alone.

```{r load-data}
fibroblasts_final <- readRDS("data/processed/kuppe_fibroblasts_final_v2.rds")
```

I evaluated cutoffs at the top and bottom quartile and at the top and bottom
third of each score's distribution. The quartile cutoff produced 7,588 High
and 7,032 Low fibroblasts under the consensus requirement, a clean contrast
but a comparatively small combined working set. The top and bottom third
cutoff produced 10,284 High and 9,864 Low fibroblasts, a meaningfully larger
set while still retaining agreement between both scoring methods, so I proceeded with it.

```{r define-groups}
aucell_high_cutoff <- quantile(fibroblasts_final@meta.data$SenMayo_AUCell, 0.67)
aucell_low_cutoff <- quantile(fibroblasts_final@meta.data$SenMayo_AUCell, 0.33)
addmod_high_cutoff <- quantile(fibroblasts_final@meta.data$SenMayo1, 0.67)
addmod_low_cutoff <- quantile(fibroblasts_final@meta.data$SenMayo1, 0.33)

fibroblasts_final@meta.data$senescence_group_33 <- "Mid"

fibroblasts_final@meta.data$senescence_group_33[
  fibroblasts_final@meta.data$SenMayo_AUCell >= aucell_high_cutoff &
  fibroblasts_final@meta.data$SenMayo1 >= addmod_high_cutoff
] <- "High"

fibroblasts_final@meta.data$senescence_group_33[
  fibroblasts_final@meta.data$SenMayo_AUCell <= aucell_low_cutoff &
  fibroblasts_final@meta.data$SenMayo1 <= addmod_low_cutoff
] <- "Low"

table(fibroblasts_final@meta.data$senescence_group_33)
```

## 2.1 Patient Representation Check

Before proceeding, I checked how evenly individual patients were represented
across the High and Low groups, since an imbalance here could let one
patient's individual biology be mistaken for a general senescence signature.

```{r patient-balance-check}
de_cells <- subset(fibroblasts_final, subset = senescence_group_33 %in% c("High", "Low"))
cell_counts <- table(de_cells@meta.data$donor_id, de_cells@meta.data$senescence_group_33)
cell_counts
```

All 20 patients were represented in both groups. One sample, P13's Low group
at 23 cells, fell below a reasonable pseudobulk stability threshold. I chose
to retain P13 given how isolated this single case was, rather than reduce my
patient count, and treat this as a documented caveat to revisit as a
sensitivity check if results appear to hinge on this one sample.

---

# 3. Pseudobulk Differential Expression

I aggregated cells into one summed expression profile per patient per group
rather than treating individual cells as independent replicates. This matters because patients, not cells, are the true independent unit of replication here, and cell counts per patient varied considerably, one patient (P15) contributed over 2,000 cells to the High group but only 69 to Low. Pseudobulk aggregation gives every patient equal statistical weight regardless of how many cells they contributed, directly addressing this imbalance.

```{r pseudobulk-aggregate}
pseudobulk <- AggregateExpression(
  de_cells,
  assays = "decontX",
  group.by = c("donor_id", "senescence_group_33"),
  return.seurat = FALSE
)

pseudobulk_counts <- pseudobulk$decontX
dim(pseudobulk_counts)
```

I used the decontX corrected assay specifically, consistent with every
analysis step since Day 3's ambient RNA correction.

```{r deseq2-setup}
sample_info <- data.frame(
  sample = colnames(pseudobulk_counts),
  donor_id = sub("_(High|Low)$", "", colnames(pseudobulk_counts)),
  group = sub("^P[0-9]+_", "", colnames(pseudobulk_counts))
)

sample_info$group <- factor(sample_info$group, levels = c("Low", "High"))
sample_info$donor_id <- factor(sample_info$donor_id)
rownames(sample_info) <- sample_info$sample

counts_rounded <- round(pseudobulk_counts)

dds <- DESeqDataSetFromMatrix(
  countData = counts_rounded,
  colData = sample_info,
  design = ~ donor_id + group
)

keep_genes <- rowSums(counts(dds)) >= 10
dds <- dds[keep_genes, ]
sum(keep_genes)
```

I used a paired design, `~ donor_id + group`, so that each patient's High
and Low fibroblasts are compared against each other's own baseline before
testing whether the group difference holds consistently across all 20
patients. This removes person to person variation unrelated to senescence
and is a more statistically powerful design than an unpaired comparison
given this dataset's structure.

```{r run-deseq2}
dds <- DESeq(dds)
res <- results(dds, contrast = c("group", "High", "Low"))
res <- res[order(res$padj), ]
summary(res)
```

Of 22,555 genes with nonzero counts, 1,255 were significantly elevated and
1,474 significantly reduced in High relative to Low fibroblasts, at
adjusted p-value < 0.1. 

```{r annotate-results}
res_df <- as.data.frame(res)
res_df$ensembl_id <- rownames(res_df)
res_df$gene_symbol <- mapIds(org.Hs.eg.db, keys = res_df$ensembl_id, keytype = "ENSEMBL", column = "SYMBOL")
```

My top individual hits were CTNNB1, JUN, PAPPA, IGF1, IGFBP4, and IGFBP7, all elevated in the High group. Four of these six converge on IGF signaling, a pathway independently documented as part of the senescence associated
secretory phenotype, an early and reassuring sign that this comparison is
capturing real senescence biology rather than noise.

---

# 4. Pathway Enrichment

I tested my significant gene list against MSigDB's Hallmark collection and
against Reactome, to identify which coordinated biological processes are
statistically over represented, rather than interpreting individual genes in
isolation.

```{r prepare-gene-list}
sig_genes <- rownames(res)[!is.na(res$padj) & res$padj < 0.1]
sig_symbols <- mapIds(org.Hs.eg.db, keys = sig_genes, keytype = "ENSEMBL", column = "SYMBOL")
sig_symbols <- unique(na.omit(sig_symbols))
length(sig_symbols)
```

```{r hallmark-enrichment}
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H")

enrichment_results <- enricher(
  gene = sig_symbols,
  TERM2GENE = hallmark_sets[, c("gs_name", "gene_symbol")]
)

as.data.frame(enrichment_results) %>% arrange(p.adjust) %>% head(10)
```

TNFA_SIGNALING_VIA_NFKB and EPITHELIAL_MESENCHYMAL_TRANSITION were by far my
strongest results, several orders of magnitude stronger than any other
pathway tested. NF-kB is well established in the literature as the master
transcriptional regulator of the SASP, so its emergence as my single
strongest enriched pathway is an independent confirmation that my
senescence scoring is capturing genuine SASP biology. EMT overlaps
substantially with the fibroblast to myofibroblast activation process
central to fibrosis, so together these two results describe senescent
fibroblasts as inflammatory, NF-kB driven cells undergoing an activation
process closely related to myofibroblast transition.

I specifically tested whether Wnt/beta-catenin signaling was enriched given
CTNNB1's position as my top individual hit.

```{r wnt-check}
all_hallmark_results <- enricher(
  gene = sig_symbols,
  TERM2GENE = hallmark_sets[, c("gs_name", "gene_symbol")],
  pvalueCutoff = 1,
  qvalueCutoff = 1
)

as.data.frame(all_hallmark_results) %>% filter(ID == "HALLMARK_WNT_BETA_CATENIN_SIGNALING")
```

Wnt/beta-catenin signaling as a coordinated pathway was not enriched
(p = 0.28, fold enrichment 1.25, close to the baseline of no enrichment).
CTNNB1 remains a genuine, strong individual hit, appearing across three
other significant pathways rather than acting in isolation.

```{r reactome-enrichment}
reactome_sets <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME")

reactome_results <- enricher(
  gene = sig_symbols,
  TERM2GENE = reactome_sets[, c("gs_name", "gene_symbol")]
)

as.data.frame(reactome_results) %>% arrange(p.adjust) %>% head(10)

write.csv(
  as.data.frame(reactome_results) %>% arrange(p.adjust),
  "results/tables/reactome_enrichment_full.csv",
  row.names = FALSE
)
```

Reactome's finer granularity resolved two pathways that Hallmark's broader
categories only showed as scattered individual genes. EXTRACELLULAR_MATRIX_
ORGANIZATION was my single strongest Reactome result, directly relevant given this is the molecular machinery of fibrosis itself. SIGNALING_BY_VEGF and the dedicated IGF transport and uptake pathway were both independently
significant, confirming VEGFA/VEGFC and the IGF/IGFBP axis observed among my
top individual hits are genuine, coordinated pathway level signals rather
than isolated genes.

---

# 5. Candidate Gene Shortlisting

I combined statistical strength with pathway convergence, counting how many
of five independently confirmed significant pathways (ECM organization, VEGF
signaling, IGF regulation, NF-kB signaling, EMT) each candidate gene belongs
to.

```{r build-candidate-table}
ecm_genes <- strsplit(as.data.frame(reactome_results)["REACTOME_EXTRACELLULAR_MATRIX_ORGANIZATION", "geneID"], "/")[[1]]
vegf_genes <- strsplit(as.data.frame(reactome_results)["REACTOME_SIGNALING_BY_VEGF", "geneID"], "/")[[1]]
igf_genes <- strsplit(as.data.frame(reactome_results)["REACTOME_REGULATION_OF_INSULIN_LIKE_GROWTH_FACTOR_IGF_TRANSPORT_AND_UPTAKE_BY_INSULIN_LIKE_GROWTH_FACTOR_BINDING_PROTEINS_IGFBPS", "geneID"], "/")[[1]]
nfkb_genes <- strsplit(as.data.frame(enrichment_results)["HALLMARK_TNFA_SIGNALING_VIA_NFKB", "geneID"], "/")[[1]]
emt_genes <- strsplit(as.data.frame(enrichment_results)["HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "geneID"], "/")[[1]]

candidates <- res_df %>%
  filter(!is.na(padj), padj < 0.01, abs(log2FoldChange) > 0.5) %>%
  arrange(padj)

candidates$in_ECM <- candidates$gene_symbol %in% ecm_genes
candidates$in_VEGF <- candidates$gene_symbol %in% vegf_genes
candidates$in_IGF <- candidates$gene_symbol %in% igf_genes
candidates$in_NFKB <- candidates$gene_symbol %in% nfkb_genes
candidates$in_EMT <- candidates$gene_symbol %in% emt_genes
candidates$pathway_count <- rowSums(candidates[, c("in_ECM", "in_VEGF", "in_IGF", "in_NFKB", "in_EMT")])

candidates <- candidates %>%
  select(gene_symbol, ensembl_id, everything()) %>%
  arrange(desc(pathway_count), padj)

rownames(candidates) <- NULL
write.csv(candidates, "results/tables/candidate_genes_full.csv", row.names = FALSE)
```

One gene in this ranked list, CCN1, connects directly to prior literature.
A 2024 Cell Death Discovery review on cardiac cell senescence reports that
CCN1, a SASP protein, ameliorates myocardial fibrosis, possibly by inducing
fibroblast senescence after MI. CCN1 independently emerged in my own top
tier with a pathway count of 3, statistically significant (padj = 3.6e-13)
and elevated in High senescence fibroblasts, which is consistent with this
published mechanism, and strengthens confidence that my scoring and comparison are recovering biologically real signal.

I took the top 15 genes by this ranking as an established, data driven
shortlist, and separately added ACVR1B and AXL, two genes that were
individually significant but showed no dedicated pathway enrichment in my
five tested pathways and comparatively limited representation in prior
senescence specific literature, as a deliberate exploratory addition.

```{r shortlist}
shortlist <- candidates %>%
  filter(pathway_count >= 2) %>%
  arrange(desc(pathway_count), padj) %>%
  head(15)

novel_additions <- res_df %>% filter(gene_symbol %in% c("ACVR1B", "AXL"))

working_list <- bind_rows(shortlist, novel_additions) %>%
  distinct(gene_symbol, .keep_all = TRUE)
```

---

# 6. Druggability and Therapeutic Evidence

I queried the Open Targets Platform for clinical stage drug and candidate
evidence, and DGIdb for broader drug-gene interaction evidence, for each of
my 17 shortlisted genes, to separate established, clinically validated
targets from genuinely novel candidates with no current therapeutic
development.

```{r opentargets-query}
final_query <- '
query($ensemblId: String!) {
  target(ensemblId: $ensemblId) {
    approvedSymbol
    drugAndClinicalCandidates {
      count
      rows {
        maxClinicalStage
        drug {
          name
        }
      }
    }
  }
}'

get_open_targets_drugs <- function(ensembl_id, symbol) {
  response <- POST(
    url = "https://api.platform.opentargets.org/api/v4/graphql",
    body = toJSON(list(query = final_query, variables = list(ensemblId = ensembl_id)), auto_unbox = TRUE),
    encode = "json",
    add_headers("Content-Type" = "application/json")
  )
  result <- content(response, as = "parsed")
  count <- result$data$target$drugAndClinicalCandidates$count
  if (is.null(count) || count == 0) {
    return(data.frame(gene_symbol = symbol, ensembl_id = ensembl_id, drug_count = 0, top_drug = NA, max_phase = NA))
  }
  top_row <- result$data$target$drugAndClinicalCandidates$rows[[1]]
  data.frame(gene_symbol = symbol, ensembl_id = ensembl_id, drug_count = count, top_drug = top_row$drug$name, max_phase = top_row$maxClinicalStage)
}

opentargets_results <- list()
for (i in 1:nrow(working_list)) {
  opentargets_results[[i]] <- get_open_targets_drugs(working_list$ensembl_id[i], working_list$gene_symbol[i])
  Sys.sleep(0.5)
}
opentargets_df <- bind_rows(opentargets_results)
```

Open Targets' `drugAndClinicalCandidates` field specifically tracks drugs
that have entered real clinical development, a narrower and more
therapeutically meaningful measure than a general interaction database.

```{r dgidb-query}
get_dgidb_drugs <- function(symbol) {
  query <- '
  query($names: [String!]) {
    genes(names: $names) {
      nodes {
        name
        interactions {
          drug { name }
          interactionScore
        }
      }
    }
  }'
  response <- POST(
    url = "https://dgidb.org/api/graphql",
    body = toJSON(list(query = query, variables = list(names = list(symbol))), auto_unbox = TRUE),
    encode = "json",
    add_headers("Content-Type" = "application/json")
  )
  result <- content(response, as = "parsed")
  nodes <- result$data$genes$nodes
  if (length(nodes) == 0 || length(nodes[[1]]$interactions) == 0) {
    return(data.frame(gene_symbol = symbol, dgidb_drug_count = 0, top_dgidb_drug = NA))
  }
  interactions <- nodes[[1]]$interactions
  top_interaction <- interactions[[which.max(sapply(interactions, function(x) {
    score <- x$interactionScore
    if (is.null(score)) -Inf else score
  }))]]
  data.frame(gene_symbol = symbol, dgidb_drug_count = length(interactions), top_dgidb_drug = top_interaction$drug$name)
}

dgidb_results <- list()
for (i in 1:nrow(working_list)) {
  dgidb_results[[i]] <- get_dgidb_drugs(working_list$gene_symbol[i])
  Sys.sleep(0.5)
}
dgidb_df <- bind_rows(dgidb_results)
```

DGIdb draws from a much broader set of sources, including preclinical
research reagents, and does not require any clinical development stage. I therefore used DGIdb's count only as a secondary, general
indicator of research attention, and used Open Targets' `drug_count` and
`max_phase` specifically to define my established versus exploratory tiers,
since that field is curated for clinical relevance.

```{r final-table}
final_candidates <- working_list %>%
  left_join(opentargets_df, by = c("gene_symbol", "ensembl_id")) %>%
  left_join(dgidb_df, by = "gene_symbol") %>%
  select(gene_symbol, log2FoldChange, padj, drug_count, max_phase, dgidb_drug_count)

final_candidates$in_ECM <- final_candidates$gene_symbol %in% ecm_genes
final_candidates$in_VEGF <- final_candidates$gene_symbol %in% vegf_genes
final_candidates$in_IGF <- final_candidates$gene_symbol %in% igf_genes
final_candidates$in_NFKB <- final_candidates$gene_symbol %in% nfkb_genes
final_candidates$in_EMT <- final_candidates$gene_symbol %in% emt_genes
final_candidates$pathway_count <- rowSums(final_candidates[, c("in_ECM", "in_VEGF", "in_IGF", "in_NFKB", "in_EMT")])

final_candidates$tier <- ifelse(
  final_candidates$drug_count > 0 & !is.na(final_candidates$drug_count),
  "Established (clinical-stage evidence)",
  "Exploratory (no clinical-stage evidence)"
)

write.csv(final_candidates, "results/tables/final_candidate_genes.csv", row.names = FALSE)

final_candidates[, c("gene_symbol", "pathway_count", "drug_count", "max_phase", "tier")]
```

This produced 7 established candidates with real clinical stage evidence
(TNC, MMP2, VEGFA, SERPINE1, CD44, FN1, IL6, VEGFC, ICAM1, AXL) and 6
exploratory candidates with strong statistical and pathway support but no
current drug development (CCN1, JUN, IGFBP4, GEM, PLAUR, THBS1, ACVR1B).
TNC in particular already has a Phase 3 antibody candidate (Tenatumomab),
and given its role in the ECM organization pathway, its convergence across 4
of my 5 tested pathways, and prior cardiac tissue remodeling literature, it
is my strongest single established candidate.

---

# 7. Summary

Comparing high and low senescence cardiac fibroblasts, defined by consensus
agreement between two independent scoring methods across 20 matched
patients, identified 2,729 significantly different genes. Pathway
enrichment converged on NF-kB driven SASP signaling, EMT associated
fibroblast activation, extracellular matrix organization, and a coordinated
VEGF and IGF signaling axis, consistent with established senescence and
fibrosis biology and with at least one direct precedent in the literature
(CCN1). From this evidence I produced a 17 gene, two tier candidate list for
anti-fibrotic gene therapy: 7 established genes with existing clinical stage
development, and 6 exploratory genes with strong statistical and pathway
support but no current therapeutic development. This analysis is
deliberately scoped to the fibroblast lineage specifically, given its direct
mechanistic role in fibrosis, with cross-cell-type signaling identified as a
next step.

---

# 8. Session Info

```{r session-info}
sessionInfo()
```
